Chromosomal duplication can be an evolutionary detour on the road to adaptation

### 2022-09-08 ms/ms.tex
add new paragraphs on new results

% Abstract 250 words
\begin{abstract} 
BLA
%Aneuploidy is common in eukaryotes, often leading to decreased cell growth and fitness. However, evidence from yeast and fungi, as well as human tumour cells, suggests that aneuploidy can be beneficial under stressful conditions and lead to elevated growth rates and adaptation.
%Importantly, aneuploidy differs from point mutations in rate, fitness effect, and reversibility.
%Here, we develop an evolutionary model for adaptive evolution with both aneuploidy and  mutation, and apply it within an Bayesian inference framework.
%We estimate the formation rate and fitness effect of aneuploidy and mutation in an evolutionary experiment in which \yeast adapted to heat stress.
%Our results suggest that aneuploidy is more frequent than previously estimated and that experimental observations cannot be explained without aneuploidy.
%We do not find evidence for genetic instability in aneuploid cells.
%Our results support the hypothesis that aneuploidy is a transient adaptive hypothesis: the experimental populations adapted by first acquiring a chromosome duplication; then by acquiring a beneficial mutation that likely appeared in aneuploid cells, rather than euploid cells; and finally by rapidly losing the extra chromosome to avoid paying the fitness cost of aneuploidy.
\end{abstract}

@@

%\paragraph{The present study.}
Here, we test the hypothesis that aneuploidy is a \emph{transient adaptive solution}.
First, we analyze previously unpublished sequencing data from the original experimental populations of \citet{Yona2012} to assess if the aneuploid population and the evolved euploid population carry the same mutant alleles, as expected if the later is descended from the former.
Second, we develop an evolutionary genetic model and fit it to the experimental results of \citet{Yona2012} in order to predict the genotype frequency dynamics in the experimental populations, thereby estimating the frequency of evolved euploid cells that descended from aneuploid cells. 
Our results show that aneuploidy reached high frequencies in the experimental populations, but nevertheless, the majority of cells in the evolved euploid populations likely did not descend from aneuploid cells, but rather directly from wild-type euploid cells.
These results suggest that aneuploidy can be an evolutionary detour, rather than a stepping stone, on the road to adaptation.

@@

% TODO F_A vs N

% TODO log-log scale of freq dynamics

@@

% Fig dynamics
\begin{figure}[h!]
      \centering
      \includegraphics[width=0.8\textwidth]{../figures/FA.pdf}      
  \caption{
    \textbf{Predicted frequency of aneuploid-descended cells.}     
    \textbf{(A)} Posterior predicted genotype frequencies over time, including the source of \eumt: $2n^*_A$ arose from $2n+1$, whereas $2n^*_M$ arose directly from $2n$. Colored curves are 50 simulations using the MAP estimate parameters. Black dashed curves are the expected genotype frequencies without genetic drift (from a deterministic model). 
     See \Cref{fig:MAP_freq_dynamics_log} for log-log scale, in which the sequence of events is easier to observe.
    \textbf{(B)} Posterior distribution of $F_A$, the expected frequency of \eumt\ cells descended from \anwt\ cells, computed as the average frequency at the end of 100 simulations for 100,000 samples from the parameter posterior distribution.
    Solid and dashed lines show the mean and 95\% CI.
	\textbf{(C)} $F_A$ values (color coded) from panel B, with their corresponding mutation rate $\mu$ on x-axis and aneuploidy rate $\delta$ on the y-axis.
	White star shows the MAP estimate. See also \Cref{fig:posterior_FA_supp}.
	\textbf{(D)} $F_A$ as a function of the population size, $N$, in posterior predictions with MAP parameters. Markers show $F_A$ in 250 simulations per population size. Error bars show mean $F_A$ with 95\% CI (bootstrap, $n=10,000$). Vertical dashed line for population size in the experiment, $6.425\cdot10^6$. Horizontal line for $F_A^{MAP}=0.106$. 
	\textbf{(E)} Expected genotype frequencies (without drift) using MAP estimate parameters. Solid lines for dynamics with aneuploidy, where \eumt\ fixates (frequency >99\%) at generation 1,900. Dashed lines for dynamics without aneuploidy ($\delta=0$), where \eumt\ fixates much earlier, at generation 500.
     }
  \label{fig:FA}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

This happens despite aneuploidy reaching a high frequency in the population (>95\%). 
Conventional wisdom might suggest that once the aneuploid genotype \anwt\ reaches high frequency, it will have a better chance at producing "refined" solutions via mutations, and its descendants will come to dominate the population: the frequency \eumtA\ (which arises from \anmt) will be higher than the frequency of \eumtM\ (which arises directly from \euwt).
So how does \eumtM\ prevail?
% TODO log-log scale of freq dynamics
At the first stage, \anwt\ cells appear and reach a high frequency due to their advantage under heat.
Accordingly, \anmt\ cells appear more frequently than \eumtM. However, the fitness advantage of a "refined" mutation that emerges in the \anmt\ genotype in hindered by the cost of aneuploidy as well as the extra waiting time needed to eliminate the extra copy of the chromosome by a chromosome loss event.
Although such chromosome loss occurs fast relative to mutation, it still leaves a time window for \eumtM\ to arise and reach high frequencies.
Indeed, we estimate that the advantage of \anmt\ over \anwt\ is similar to the advantage of \eumt\ over \anmt (about 0.3\%), and that the rate of chromosome loss is roughly $1-3\cdot10^{-3}$.

% TODO effect of population size?

@@

\paragraph{Conclusions.}
Here, we tested the hypothesis that aneuploid cells are an evolutionary adaptive intermediate between wild-type euploid cells and mutant euploid cells. 
Our results suggest this is only true at the population level, as the population does go from euploid to aneuploid and back--but not at the individual level: we estimate that only about 10-15\% of the euploid cells descended from aneuploid cells, whereas the rest are direct descendent of the wild-type euploid cells.

>>>>

% Abstract 250 words
\begin{abstract} 
BLA % TODO revise
%Aneuploidy is common in eukaryotes, often leading to decreased cell growth and fitness. However, evidence from yeast and fungi, as well as human tumour cells, suggests that aneuploidy can be beneficial under stressful conditions and lead to elevated growth rates and adaptation.
%Importantly, aneuploidy differs from point mutations in rate, fitness effect, and reversibility.
%Here, we develop an evolutionary model for adaptive evolution with both aneuploidy and  mutation, and apply it within an Bayesian inference framework.
%We estimate the formation rate and fitness effect of aneuploidy and mutation in an evolutionary experiment in which \yeast adapted to heat stress.
%Our results suggest that aneuploidy is more frequent than previously estimated and that experimental observations cannot be explained without aneuploidy.
%We do not find evidence for genetic instability in aneuploid cells.
%Our results support the hypothesis that aneuploidy is a transient adaptive hypothesis: the experimental populations adapted by first acquiring a chromosome duplication; then by acquiring a beneficial mutation that likely appeared in aneuploid cells, rather than euploid cells; and finally by rapidly losing the extra chromosome to avoid paying the fitness cost of aneuploidy.
\end{abstract}

@@

%\paragraph{The present study.}
% TODO revise
Here, we test the hypothesis that aneuploidy is a \emph{transient adaptive solution}.
First, we analyze previously unpublished sequencing data from the original experimental populations of \citet{Yona2012} to assess if the aneuploid population and the evolved euploid population carry the same mutant alleles, as expected if the later is descended from the former.
Second, we develop an evolutionary genetic model and fit it to the experimental results of \citet{Yona2012} in order to predict the genotype frequency dynamics in the experimental populations, thereby estimating the frequency of evolved euploid cells that descended from aneuploid cells. 
Our results show that aneuploidy reached high frequencies in the experimental populations, but nevertheless, the majority of cells in the evolved euploid populations likely did not descend from aneuploid cells, but rather directly from wild-type euploid cells.
These results suggest that aneuploidy can be an evolutionary detour, rather than a stepping stone, on the road to adaptation.

@@

@@

% Fig dynamics
\begin{figure}[h!]
      \centering
      \includegraphics[width=0.8\textwidth]{../figures/FA.pdf}      
  \caption{
    \textbf{Predicted frequency of aneuploid-descended cells.}     
    \textbf{(A)} Posterior predicted genotype frequencies over time, including the source of \eumt: $2n^*_A$ arose from $2n+1$, whereas $2n^*_M$ arose directly from $2n$. Colored curves are 50 simulations using the MAP estimate parameters. Black dashed curves are the expected genotype frequencies without genetic drift (from a deterministic model). 
     See \Cref{fig:MAP_freq_dynamics_log} for log-log scale, in which the sequence of events is easier to observe.
    \textbf{(B)} Posterior distribution of $F_A$, the expected frequency of \eumt\ cells descended from \anwt\ cells, computed as the average frequency at the end of 100 simulations for 100,000 samples from the parameter posterior distribution.
    Solid and dashed lines show the mean and 95\% CI.
	\textbf{(C)} $F_A$ values (color coded) from panel B, with their corresponding mutation rate $\mu$ on x-axis and aneuploidy rate $\delta$ on the y-axis.
	White star shows the MAP estimate. See also \Cref{fig:posterior_FA_supp}.
	\textbf{(D)} $F_A$ as a function of the population size, $N$, in posterior predictions with MAP parameters. Markers show $F_A$ in 250 simulations per population size. Error bars show mean $F_A$ with 95\% CI (bootstrap, $n=10,000$). Vertical dashed line for population size in the experiment, $6.425\cdot10^6$. Horizontal line for $F_A^{MAP}=0.106$. 
	\textbf{(E)} Population mean fitness in a model without drift using MAP estimate parameters. Solid lines for mean fitness with aneuploidy ($\delta>0$), where the population reaches adaptation (mean fitness at 99.99\% of maximum value) at generation 1,802. Dashed lines for mean fitness without aneuploidy ($\delta=0$), where the population adapts much earlier, at generation 535.
     }
  \label{fig:FA}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

This happens despite aneuploidy reaching a high frequency in the population (>95\%). 
Conventional wisdom might suggest that once the aneuploid genotype \anwt\ reaches high frequency, it will have a better chance at producing "refined" solutions via mutations, and its descendants will come to dominate the population: the frequency \eumtA\ (which arises from \anmt) will be higher than the frequency of \eumtM\ (which arises directly from \euwt).

So how does \eumtM\ prevail?
The supply rates of \anwt\ and \eumtM\ are $N \delta \approx 11,000$ and $N \mu \approx 19$, respectively (assuming MAP parameter estimates). 
Therefore, both genotypes are expected to appear immediately at the beginning of the experiment (\Cref{fig:MAP_freq_dynamics_log}).
However, \anwt\ appears at a much higher frequency as $\delta \gg \mu$ by 2-3 orders of magnitude.
After they first appear, \eumtM\ has higher fitness.
But as long as the frequency of \euwt\ is high, the supply rate of \anwt\ is higher than that of \eumtM, again due to $\delta \gg \mu$. 
However, because both genotypes are produced from \euwt.
Therefore, when the frequency of \euwt\ decreases, mainly due to the increase in the frequency of \anwt, both supply rates diminish.
At this point, the higher fitness of \eumtM\ comes into play and it starts to take over the population, which at this stage is mainly composed of \anwt.
For the aneuploid lineage to compete with the mutant lineage, it must produce \eumtA\ via a mutation followed by chromosome loss.
Although this is a stochastic process (due to drift), our results show that the time until \eumtA\ reaches a frequency of ~0.001 is roughly 450 generation, without a lot of variation (intersection of purple lines and vertical dashed line in \Cref{fig:MAP_freq_dynamics_log}).
However, by that time, \eumtM\ is already at a much higher frequency, ~0.18, and since both mutants have the same fitness, their relative frequency remains roughly the same until the end of the experiment. 
%At the first stage, \anwt\ cells appear and reach a high frequency due to their advantage under heat.
%Accordingly, \anmt\ cells appear more frequently than \eumtM. However, the fitness advantage of a "refined" mutation that emerges in the \anmt\ genotype in hindered by the cost of aneuploidy as well as the extra waiting time needed to eliminate the extra copy of the chromosome by a chromosome loss event.
%Although such chromosome loss occurs fast relative to mutation, it still leaves a time window for \eumtM\ to arise and reach high frequencies.
%Indeed, we estimate that the advantage of \anmt\ over \anwt\ is similar to the advantage of \eumt\ over \anmt (about 0.3\%), and that the rate of chromosome loss is roughly $1-3\cdot10^{-3}$.

% F_A vs N
\paragraph{Prediction for a small populations.}
We examined the effect of the population size, $N$, on the frequency of \anwt\ descendants in the evolved population, $F_A$.
We found that $F_A$ is expected to decrease as the population size increases (\Cref{fig:FA}D), ranging from about 0.9 when the population size is 10,000, to about 0.1 when the population size is above 1,000,000 (less than the experimental population size, which was 6,425,000). 
Thus, our model provides a testable prediction: if the experiment was repeated under a lower population size (via stronger daily dilutions or with in a smaller volume), then the fraction of the population descending from aneuploid cells would likely be much higher.

% mean fitness
\paragraph{Aneuploidy increases adaptation time.}
We found that aneuploidy quickly appears and reaches high frequencies but is then outcompeted by adaptive mutant lineages. 
An additional interesting outcome is that aneuploidy increases rather than decreases the adaptation time (expected time until the population mean fitness is at 99.99\% of its maximum value, which is 1,802 and 535 generation with and without aneuploidy, respectively; \Cref{fig:FA}E). 
This happens despite the fact that the mean fitness initially increases faster in the presence of aneuploidy (\Cref{fig:FA}E). 
The reason is that once \anwt\ is common, selection for the mutant strain (\anmt\ or \eumt) is weaker compared to when \eumt\ competes directly with \euwt.

@@

\paragraph{Conclusions.}
% TODO revise
Here, we tested the hypothesis that aneuploid cells are an evolutionary adaptive intermediate between wild-type euploid cells and mutant euploid cells. 
Our results suggest this is only true at the population level, as the population does go from euploid to aneuploid and back--but not at the individual level: we estimate that only about 10-15\% of the euploid cells descended from aneuploid cells, whereas the rest are direct descendent of the wild-type euploid cells.
